Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-08-07 manuscript.rmd
Incorporated Chonggang's comment; first run

```{r, include=FALSE, echo = FALSE}
if(!require("pacman", quietly=TRUE))install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom))
options(tinytex.verbose = TRUE)
source("code/load.R")
source("code/manuscript_data_figs.R")
# load("results/manuWorkSpace.RData")
# install.packages(c("vctrs", "rlang"))
```

@@

- Deep water access is believed to be an important, drought-tolerance mechanism, but under-studied given the lack of data. Vulnerability to embolism ($\Psi_{88,~stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test the role of deep-water access in drought-tolerance strategies of a hyper-diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$) that scaled new data on 21 species with Leaf Mass Area~lamina~ and Wood Specific Gravity~100^$\circ$^C~. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, but lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for predicting the future of tropical forests under climate change.

@@

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size. Consequently trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the World’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 throusand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground. Such data are available only for ~0.05% of World's tree species [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests. Further, to estimate plant available water with soil depth requires hydrological data and modeling, but these are rarely co-located with forest inventory sites (but see, @Johnson:2018fad; @Kumagai:2011; @Ivanov:2012)

\
Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied. Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions). A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Johnson:2018fad]. A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015]. @Chitra-Tarak.2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts. They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum. Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration. However, lack of data on rooting or water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.

\
Trade-offs shape how trees' make a living in their environment (ecological strategies) and also provide Dynamic Vegetation Models (DVMs) the essential constraints to represent species-rich communities under changing environment [@Koven:2020]. This entails indentifying trade-offs among environment, traits and demography. @Reich:2014 concenptualised a universal 'fast-slow' (aka productive-persistent, or acquisitive-conservative) plant economics spectrum coordinated for all resources and organs, both above- and below-ground. specifically for water observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified a fast-slow spectrum, although examples of all trade-offs observed simultaneously are rare, and belowground trait or environment linkages are weak. 'Fast' species at the productive end of the spectrum have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with slow traits are better protected from C losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Fast strategy is associated with spatially wet sites such as valleys or high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the fast-slow spectrum holds true for the moisture dynamics vertically belowground is not clear. Deep rootedness may be expected to support a fast strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In contrast, in pulse-driven arid systems *shallow*-rooted plants are associated with fast traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show slow traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

\
Here, for large trees of length(erd.sp) canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 34 years that withnessed droughts of various intensity and duration. We obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for xx species. Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community. We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

\
We test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on a fast-slow economic spectrum optimised for increasingly reliable water availability by depth such that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

@@

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD; 1983-2017) and a four-month long dry season from mid-December through mid-April (≤ 100 mm rain/month). The entire island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

\
In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere.

@@

\
To estimate dynamics of water availability by depth over the forest’s root zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). FATES is a state-of-the-art, community-based, open-source model used for studying climate-vegetation interactions. FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach...... (How ELM talks to FATES)...Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].

@@

\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig. \@ref(fig:hydro-calib); daily, 2015-2018). We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm). TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig. \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^). The same calibration equation was used for other probes. We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019). Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables.) See Table \@ref(tab:param) for best-fit parameter ranges. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

@@

## Model for Leaf Vulnerability Curves at Community-scale
\
We measured ($K_{leaf}$ vs. $\Psi_{leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI x). We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011. Exponential form was chosen as it best-fitted all but one species based on AIC criteria (Fig. S\@ref(fig:obs-plc-curves)), described as

@@

where $A$ and $B$ are fitted species-specific parameters.

We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ (or, WSG~100^$\circ$^C~) and $LMA$ (or, LMA~lamina~), defined, respectively, as mean wood specific gravity after drying at 100$^\circ$C (gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (gm^-2^) for leaves receiving direct sunlight.  Fitted polynomial equations are described as,

@@

where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients. $WSG$ & parameter $B$ explained large variation in parameter $A$ (Eq. \@ref(eq:a-mod); Adj. R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig. S\@ref(fig:ab-fit)). $WSG$ & $LMA$ explained large variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`). We thus sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig. \@ref(fig:comm-plc-curves)).

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
# knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
```

@@

Water is a key limiting resource for tree growth. Vapour pressure deficit (VPD) and species-specific maximum hydraulic conductivity and its decline with water-stress ($K_{leaf}$ vs. $\Psi_{soil}$), are factors known to influence trees water uptake [add references, Chonggang?]. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)- \@ref(eq:M4)), in which a species growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as:

>>>>

```{r, include=FALSE, echo = FALSE}
if(!require("pacman", quietly=TRUE)) install.packages("pacman")
suppressPackageStartupMessages(
  pacman::p_load(bookdown, knitr, pander, citr, broom))
options(tinytex.verbose = TRUE)
source("code/load.R")
source("code/manuscript_data_figs.R")
# load("results/manuWorkSpace.RData")
# install.packages(c("vctrs", "rlang"))
```

@@

- Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. Vulnerability to embolism ($\Psi_{88,~stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. 
- We tested the role of deep-water access in survival strategies of a diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach. We inversely modelled effective rooting depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We evaluated our ERD estimates with existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, and lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that plants mitigate mortality risk from water stress through deep water access.
- The role of deep water access in alleviating mortality of drought-sensitive trees has important implications for our predictive understanding of tropical forest dynamics under current and future climate.

@@

# Introduction
\
Tropical forests play a disproportionately large role in the global carbon and energy cycle, and support half of global biodiversity [@Bonan:2008; @Wright:20050th], but face a threat from intensifying droughts [@Doughty:2015; @Xu.2019]. Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]. Historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; but see @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood or quantified.

\
Among all the mechanisms plants employ to deal with drought, deep water access is arguably the most commonly invoked, but under-studied mechanism. This mechanism assumes that water availability varies belowground by vertical depth and so do rooting depths of co-occuring trees, by species and size [@Meinzer:1999; @Chitra-Tarak.2018]. Consequently, trees' in the same forest differ in their drought experience, and thus in their responses in term of growth and mortality. A key bottlneck in community-wide testing of this mechanism has been lack of data on trees' rooting or water-sourcing depths, followed by data on water-availability at those depths. A recent meta-analysis documented maximum rooting depths for 318 tree species, that is, less than 0.5% of the world’s 65 thousand tree species [@Fan:2017; @Beech:2017]. Further, only a small fraction of those are tropical, even though ~60-80 % (40 - 53 thousand species) of the World’s tree diversity resides in the tropics [@Slik:2015]. Use of stable isotopes of water as a tracer provides an indirect way of trees' water sourcing depths by matching isotopic value in plant tissue to variations in soil pore water vertically belowground. However, such data are extremely rare [@Evaristo:2016]. In general, community-scale data collection for rooting or water-sourcing depths is especially challenging for species-rich tropical forests.

\
Trade-offs among co-occurring plants shape how they make a living in their environment (ecological strategies) and also provide the essential constraints to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d]. This entails indentifying trade-offs among environment, traits and demography. Specifically for the water environment (or resource) observations of individual trade-offs between environment-trait, trait-trait and trait-demography has identified an acquisitive-conservative resource economic spectrum similar to the ones for light and nutrients, although, for water, examples of all trade-offs observed simultaneously are rare, and belowground traits or environment linkages are weak (@Reich:2014; @Sterck:2011). Acquisitive species have traits with high capacity to acquire, move, store and use water (high $K_{max}$, low wood density) and achieve high growth potential, but build flimsy, disposable tissues (high SLA) and are less tolerant of low resources, whereas species with conservative traits are better protected from carbon losses and are more drought tolerant (low $\Psi_{50}$, low $\Psi_{tlp}$, larger safety margin). Acquisitive species are spatially associated with wet sites (high resource) such as valleys or those with high rainfall [@Kursar:2009; @Oliveira:2019], but whether and how the acquisitive-conservative spectrum holds true for the moisture dynamics vertically belowground is not clear. Deep rootedness may be expected to support an acquisitive strategy with a steady water supply unlike shallow-rooted trees. However, this may depend on the exact dynamics of water-availability by depth at a site, depending on precipitation regime and hydrology. In contrast, in pulse-driven arid systems *shallow*-rooted plants are associated with acquisitive traits (high $K_{max}$) that maximise the use of the water-pulse, while *deep*-rooted plants show conservative traits (cavitation-resistant xylem, osmoregulation) [@Schwinning:2013].

\
Huge efforts are invested in linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating that plants reach critical thresholds of hydraulic failure depending on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions). However, plants' realised water environments are rarely studied, especially in the tropical region. To better understand how belowground rooting is linked to aboveground hydraulic traits and mortality for co-occuring tropical forest tree species, we make a thorough assessment of trade-offs between the belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 34 years that witnessed droughts of various intensity and duration for a tropical site. Specifically, we test an overarching hypothesis that interspecific variation in effective rooting depths of trees aligns species on an acquisitive-conservative economic spectrum optimised for increasingly reliable water availability by depth such that (i) trees with deeper ERD have hydraulic traits associated with rapid water transport but cavitation-vulnerable xylem: higher maximum stem hydraulic conductance ($K_{max}$), but greater xylem vulnerability to embolism ($\Psi_{88,~stem}$), higher $\Psi_{tlp}$ and narrower hydraulic safety margins ($\Psi_{min} - \Psi_{88,~stem}$), whereas shallower ERD species are associated with less efficient but safer hydraulic traits; and that, (ii) deeper ERD species have lower mortality rates due to low realised risk of hydraulic failure, whereas shallower ERD species may have higher mortality rates as a result of higher variability in shallow soil water availability. To our knowledge this is the first study to demonstrate a mechanistic link between trees' realised drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a species-rich forest.

@@

In summary, we obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for 21 species. Identifying scaling relationships between these curves and commonly available traits (Leaf Mass Area~lamina~, Wood Specific Gravity~100^$\circ$^C~; defined below) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community (of ~ 320 total). We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

## Study Site Description
\
We conducted this study at the seasonally dry tropical forest on Barro Colorado Island (BCI), Panama, given its extensive characterization in terms of long-term forest demography, plant functional ecology, hydrology, and water sourcing depth water isotopes on tree species and soil from an independent isotopic tracer study on trees’ water-sourcing depths [@Meinzer:1999]. The 1,560ha island was formed after the river Chagres was dammed in 1914 that created Gatun Lake, the main channel of the Panama Canal. Rainfall at BCI is seasonal with a mean annual rainfall of 2687 mm (± 515 SD; 1983-2017) and a four-month long dry season from mid-December through mid-April ($\le$ 100 mm rain/month). The entire island is forested and classified as a tropical moist forest. Although percentage of canopy species that are at least partly deciduous during the dry season is 28%, the maximum percentage of deciduous individuals in the canopy is only 6.3%, and the percentage of crown area that is deciduous is estimated at 9.7% [@Condit:2000]. Long-term hydrological monitoring at BCI began in 1972, while demographical monitoring began with the 50ha ForestGEO plot establishment in 1981-82 [@Condit.1998uq; @Hubbell:1999; @Anderson-Teixeira:2014].

\
In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100-year-old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere[xx add references].

@@

\
To estimate dynamics of water availability by depth over the forest’s root zone, we implemented a 1-D hydrological model, the Energy Exascale Earth System Land Model (ELM), coupled with a Functionally Assembled Terrestrial Ecosystem Simulator (FATES) (hereafter, ELM-FATES). FATES is a state-of-the-art, community-based, open-source model used for studying climate-vegetation interactions. FATES is considered a next generation vegetation dynamics model for ELM, with a size-structured group of plants (cohorts) and successional trajectory-based patches using the ecosystem demography approach. FATES couples to ELM by a common interface of water and carbon fluxes. Detailed descriptions can be found elsewhere [@Koven:2020; @Fisher:2014; @Bonan:2013].

@@

\
We calibrated the model over 2012-2018 against three key fluxes in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET, $E_{v} + E_{g}$) from the eddy flux tower (monthly, 2012-2017), (ii) stream discharge for the Conrad Catchment ($q_{over} + q_{drai}$; daily, 2012-2018) and (iii) volumetric water content (VWC; $W_{liq, z}$; Fig. \@ref(fig:hydro-calib); daily, 2015-2018). We collected VWC data (2016-2018) from a point location at the eddy flux tower with Time Domain Reflectometry (TDR) using three horizontal probes at depths 10, 40 and 100 cm and one vertical probe (0-15 cm). TDR data for the vertical probe was calibrated based on measurements of gravimetric soil moisture on several occasions (Fig. \@ref(fig:TDR-calib)) and mean bulk dry density of 0.8 gcm^-3^ (*n* = 5, cylinder volume = 1250 cm^3^). The same calibration equation was used for other probes. We also leveraged existing opportunistic dry season VWC measurements for the entire 50-ha plot (15, 40 and 100 cm in 2015 and 15 & 40 cm in 2016; @Kupers:2019). Among the 5000 parameter ensembles we chose 100 ensembles that best-fit all three fluxes simultaneously (add details...that is those that maximised the sum of regression coefficients between observed and simulated variables. xx add equation) See Table \@ref(tab:param) for best-fit values and ranges for different parameters. We ran ELM-FATES with the chosen 100 ensembles from 1985-2018, with first 5 years used for model spin-up.

@@

## Model for Leaf Vulnerability Curves at Community-scale
\
We measured ($K_{leaf}$ vs. $\Psi_{leaf}$) on adult trees of `r length(unique(data.model.AB[!is.na(data.model.AB$data.A),]$sp))` common species (see SI--Joseph please add a section). We fitted exponential, sigmoidal, logistic equations to species-level data as prescribed by @Guyot:2011. Exponential form was chosen, because based on AIC criteria it best-fitted all but one species, and for the one species it fitted only slightly worse (Fig. S\@ref(fig:obs-plc-curves); we might want to add specie-specific panel of graphs showing data and fitted curves here), described as

@@

where $A$ and $B$ are fitted species-specific parameters. We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits that were available at community-scale, namely, mean wood specific gravity after drying at 100$^\circ$C ($WSG$ or, WSG~100^$\circ$^C~; gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules ($LMA$ or LMA~lamina~; gm^-2^) for leaves receiving direct sunlight. Fitted polynomial equations are described as,

@@

where $(a_0, a_1, a_2, a_3, b_0, b_1, b_2, b_3)$ are coefficients. $WSG$ & parameter $B$ explained a large proportion of variation in parameter $A$ (Eq. \@ref(eq:a-mod); Adj. R^2^ = `r signif(glance(k_by_psi.models[[2]])$adj.r.squared, 2)`; *P* = `r signif(glance(k_by_psi.models[[2]])$p.value, 2)`; also see Fig. S\@ref(fig:ab-fit)). $WSG$ & $LMA$ explained a large proportion of variation in parameter $B$ (Eq. \@ref(eq:b-mod); Adj. *R*^2^ = `r signif(glance(k_by_psi.models[[1]])$adj.r.squared, 2)`; *P* = `r  signif(glance(k_by_psi.models[[1]])$p.value, 2)`). We thus sequentially used Eq. \@ref(eq:b-mod) and \@ref(eq:a-mod) to predict parameters $B$ and $A$, respectively, for the `r length(unique(data.model.AB$sp))` species for which WSG and LMA data were available and fitted vulnerability curves (see Fig. \@ref(fig:comm-plc-curves)).

```{r comm-plc-curves, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "30%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-comm-plc-curves} Modelled standardized 'Vulnerability curves' ($K_{leaf}$ vs. $\\Psi_{leaf}$) at community-scale (", length(data.model.AB$sp)," species) based on data on $WSG$ & $LMA$ and their relationships with vulnerability curve parameters measured on ", sp.soft.filled$AB.data.model," species. Each line represents a species, color coded by $WSG$ with blue indicating a smaller value of WGS and red indicating a larger value of WSG.[xx chagne legend to a scale not just two values.]")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_SG100C_AVG_predicted_AB.jpeg"))
# knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/kmax_by_psi/Leaf/kmax_by_psi_color_by_LMALAM_AVD_predicted_AB.jpeg"))
```

@@

Water is a key limiting resource for tree growth. Vapour pressure deficit (VPD) and water supply as determined by soil moisture are factors known to influence trees water uptake [@Lawrence.2019]. In this study, we used leaf conductivity ($K_leaf$) determined by rhizosphere soil water potential ($\Psi_{soil}$) as an approximation of the soil water limiting factor. For dry-deciduous species seasonality in leaf cover may also influence plant water uptake. We therefore tested alternate models of growth, aka inverse models for ERD (Eqs. \@ref(eq:M2)- \@ref(eq:M4)), in which a species' growth dynamics across census intervals was regressed against census interval means of multiplied effects of one or more factors calculated daily: VPD, $K_{leaf}$ and the fraction of leaf cover (all standardised to 0-1, see below). Let n~t~ be the total number of days in census interval $t$, then for individual tree species Growth~t~ and depth $z$ these models are described as follows (xxadd additive models):
